Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-06 submission/manuscript.Rmd
began discussion edits

In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala. In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. Given the majority of tissue-enriched DEGs are testis and ovary genes with no known brain functions, it is currently unknown if they impair _Kdm5c_-KO brain function and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression. However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus. Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010]. Thus, failure to fine-tune the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_ could be one route by which loss of brain tissue identity contributes to _Kdm5c_-KO impairments.

We found a substantial portion of genes upregulated in _Kdm5c_-KO brain are germline genes with no known fucntions within the brain. Distinguishing the germline (i.e. cells that pass on their genetic material to the next generation) from the soma (i.e. all other cells of the body) is a key feature of multicellularity and sexual reproduction. Previous work characterizing chromatin regualtors that silence germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. To globally characterize KDM5C's role in germline gene repression throughout life, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015]. This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) the categories of germline genes directly and indirectly regulated by KDM5C.

RNA-seq of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis. _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[@bardotMouseGastrulationCoordination2020]. However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011]. Therefore, misexpression of germline genes in EpiLCs might suggest _Kdm5c_-KO ESCs are progressing beyond EpiLC differentiation and instead becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs properly express primed pluripotency genes, including _Otx2_ - an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018]. Furthermore, we observed no difference in cellular morphology during _Kdm5c_-KO ESC to EpiLC differentiation. Proper EpiLC differentiation, together with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.

We then globally characterized KDM5C binding at germline-enriched gene promoters through KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs). While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs. This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress two aberrant germline transcripts[@scandagliaLossKdm5cCauses2017]. Although enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of the germline genes aberrantly transcribed in _Kdm5c_-KO EpiLCs. One notable EpiLC DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells to promote meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. However, retinoic acid can only activate _Stra8_ expression when DAZL is present. Unlike _Stra8_, _Dazl_ is a direct target of KDM5C in EpiCs and is both transcribed and translated in _Kdm5c_-KO EpiLCs. Expression of indirect DEGs like _Stra8_ indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells through downstream activation by other ectopic germline programs. These ectopic germline programs appear to loosely mimic the trajectory of typical germline development, as germline genes important for early germ cell development and meiosis are expressed in _Kdm5c_-KO EpiLC while late spermatogenesis genes are expressed in the mature _Kdm5c_-KO brain. Altogether, this indicates once a subset of germline genes are activated due to loss of direct KDM5C repression in the early embryo, these ectopic germline programs can continue to progress in the background of _Kdm5c_-KO somatic development.

@@

>>>>

In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala. In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. Deeper characterization of testis-enriched DEGs revealed they were not somatic testis genes, but germline genes, thus demonstrating KDM5C's crucial role in establishing the soma-germline boundary. Given the majority of tissue-enriched DEGs are testis and ovary genes with no known brain functions, it is currently unknown if they impair _Kdm5c_-KO brain function and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression. However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus. Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010].

We found a substantial portion of genes upregulated in _Kdm5c_-KO brain are germline genes with no known fucntions within the brain. Distinguishing the germline from the soma is a key feature of multicellularity and sexual reproduction[@devlinPanmetazoanViewGermlinesoma2023]. Previous work characterizing chromatin regualtors that silence germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. To characterize KDM5C's role in germline gene repression at a genome-wide level throughout life, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015]. This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) which groups of germline genes are directly and indirectly regulated by KDM5C. This list of germline-enriched genes serves as a resource for future studies to systematically assess soma-germline dysregulation.

RNA-seq of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis. _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020]. However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011]. Therefore, misexpression of germline genes in EpiLCs might suggest _Kdm5c_-KO ESCs are progressing beyond EpiLC differentiation and instead becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs properly express primed pluripotency genes, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Furthermore, we observed no difference in cellular morphology during _Kdm5c_-KO ESC to EpiLC differentiation. Proper EpiLC differentiation, together with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.

We then globally characterized KDM5C binding at germline-enriched gene promoters through KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs).

Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which germline genes can be misexpressed in _Kdm5c_-KO cells.

While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs. This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress two aberrant germline transcripts[@scandagliaLossKdm5cCauses2017]. Although enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of the germline genes aberrantly transcribed in _Kdm5c_-KO EpiLCs. One notable EpiLC DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells to promote meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. However, retinoic acid can only activate _Stra8_ expression when DAZL is present. Unlike _Stra8_, _Dazl_ is a direct target of KDM5C in EpiCs and is both transcribed and translated in _Kdm5c_-KO EpiLCs. Expression of indirect DEGs like _Stra8_ indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells through downstream activation by other ectopic germline programs. These ectopic germline programs appear to loosely mimic the trajectory of typical germline development, as germline genes important for early germ cell development and meiosis are expressed in _Kdm5c_-KO EpiLC while late spermatogenesis genes are expressed in the mature _Kdm5c_-KO brain. Altogether, this indicates once a subset of germline genes are activated due to loss of direct KDM5C repression in the early embryo, these ectopic germline programs can continue to progress in the background of _Kdm5c_-KO somatic development.

@@

- Last paragraph
	- Thus, failure to fine-tune the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_ could be one route by which loss of brain tissue identity contributes to _Kdm5c_-KO impairments.
